---
title: "Mental Health and Musical Preferences"
author: "Meghan Harris"
format: html
execute:
  echo: true
  warning: false
  error: true
engine: knitr
code-fold: true
# Extension configuration
filters:
  - collapse-output
extensions:
  collapse-output:
    method: javascript
format-options:
  python:
    code-formatter: none
    path: ".venv/Scripts/python.exe"
---

# Introduction

I haven't done statistical modeling since 2021, and back then I tried using base R's `lm()` and hated it. I didn't know tidymodels existed at the time, so I assumed modeling just wasn't for me. Fast forward to 2024: I'm pivoting from data science to open source development, learning Python alongside R, and decided to revisit modeling with modern tools. This notebook compares a simple ML workflow in both R (tidymodels) and Python (scikit-learn) to:

1. Refresh my modeling skills using tidymodels (which actually makes sense to my brain)
2. Practice Python by replicating a workflow I understand in R

I'm using the [Music & Mental Health Survey Results](https://www.kaggle.com/datasets/catherinerasgaitis/mxmh-survey-results) dataset from Kaggle to predict anxiety scores from music listening habits.


# Data Import and Viewing
First, we import the data.

::: {.panel-tabset}

## R

```{r, r-data-import}
# install.packages(c("janitor", "tidyverse", "tidymodels"))
library(tidymodels)
library(tidyverse)
library(reticulate)

# Grab relative data path
data_path <-
  here::here(
    "mental_health_music",
    "data",
    "mxmh_survey_results.csv"
  )

# Import data and clean vars
df_mh_music_r <-
  read_csv(data_path) |>
  janitor::clean_names()

```

## Python

```{python, py-data-import}
import polars as pl
import janitor.polars
from pathlib import Path
from plotnine import (
    ggplot,
    aes,
    after_stat,
    element_blank,
    element_text,
    geom_abline,
    geom_bar,
    geom_point,
    geom_text,
    labs,
    scale_fill_gradient2,
    scale_x_discrete,
    theme,
    theme_minimal,
)
import textwrap

# Python has been SUPER annoying with paths for me in quarto docs
# Maybe it's me/my machine but, meh
# Setting two possible locations for the data file
possible_paths = [
    Path("data") / "mxmh_survey_results.csv",
    Path("mental_health_music") / "data" / "mxmh_survey_results.csv",
]

for path in possible_paths:
    if path.exists():
        data_path = path
        break
else:
    raise FileNotFoundError("mxmh_survey_results.csv not found in expected locations.")

# Read in csv and clean up the names
df_mh_music_py = pl.read_csv(
    data_path,
    # This is needed because there's decimals in the
    # scoring cols where integer was inferred
    infer_schema_length=1000,
).clean_names(remove_special=True)

```

:::

Then, do a little skimming with glimpse...

::: {.panel-tabset collapse=true}

## R

```{r, r-data-glimpse}
#| output-fold: true
# Take a peek
df_mh_music_r |>
  glimpse()

```

## Python

```{python, py-data-glimpse}
# | output-fold: true
# Take a peek
df_mh_music_py.glimpse()
```

:::

# Data Exploration

Looking at the data, I'm interested in predicting "anxiety score" from musical listening patterns with respect to genre frequencies.
We'll need the `anxiety` and the `frequency_*`, so let's visualize what we have here.

For the anxiety score interpretation on a scale from 0 to 10, respondents where asked "how often do you experience this?":
  
  * 0 - I do not experience this.

  * 10 - I experience this regularly, constantly/or to an extreme.

::: {.panel-tabset collapse=true}

## R

So, of no surprise to me, this data is right-skewed with a concentration sitting around a score from 7 - 8.
The density curve also shows limited variability at lower anxiety levels ("flattening" of the curve), which may affect model performance in predicting across the full range of the outcome.

```{r, r-eda-1, fig.align="center"}
df_mh_music_r |>
  ggplot(aes(x = anxiety)) +
  geom_histogram(
    aes(y = after_stat(density), fill = after_stat(count)),
    bins = 11,
    color = "#000000"
  ) +
  geom_density(color = "#000000", linewidth = 3) +
  geom_density(color = "#60bcde", linewidth = 1) +
  scale_fill_gradient2(
    low = "#ffffff",
    high = "#084164"
  ) +
  labs(
    title = "Distribution of Self-Reported Anxiety Scores",
    subtitle = "Among Survey Respondents",
    fill = "Total Responses:",
    x = "Reported Anxiety Score",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    text = element_text(size = 10),
    legend.position = "top"
  )
```

I also should look at the music genre categories as there are `r df_mh_music_r |> select(starts_with("frequency")) |> length()` different genres that can each have one of four different responses: “Never”, “Rarely”, “Sometimes”, or “Very frequently”

```{r, r-eda-2, fig.align="center"}
# Need to pivot the genres
df_genre_pivot_r <-
  df_mh_music_r |>
  pivot_longer(
    cols = starts_with("frequency")
  ) |>
  mutate(
    name_cleaned = str_remove(name, "^frequency_"),
    name_cleaned = case_when(
      name_cleaned == "k_pop" ~ "K-pop",
      name_cleaned == "r_b" ~ "r & b",
      name_cleaned == "hip_hop" ~ "hip-hop",
      name_cleaned == "edm" ~ "EDM",
      .default = str_replace_all(name_cleaned, "_", " ")
    ),
    name_cleaned = if_else(
      name_cleaned == "EDM",
      name_cleaned,
      str_to_title(name_cleaned)
    )
  )

df_genre_pivot_r |>
  ggplot(aes(x = name_cleaned, fill = value)) +
  geom_bar(
    color = "#000000"
  ) +
  scale_fill_manual(
    values = c(
      "Never" = "#FFFFFF",
      "Rarely" = "#ACBFCB",
      "Sometimes" = "#5A8097",
      "Very frequently" = "#084164"
    )
  ) +
  labs(
    title = "Distribution of Music Genre Listening Frequency",
    subtitle = "Among Survey Respondents",
    fill = "Response:",
    x = "Musical Genre",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_line(color = "#000000"),
    plot.title = element_text(size = 14, face = "bold"),
    text = element_text(size = 10),
    axis.text.x.bottom = element_text(
      size = 10,
      angle = 45,
      vjust = 1,
      hjust = 1
    ),
    legend.position = "top"
  )

```

So a lot of responses have a value of "Never" or "Rarely". Have 16 different genres with each having one of four possible responses means that the feature space is way too big because that's a possibility of 4^`16 combinations (`r scales::comma(4^16)`) !!!

I'm going to make a binary response variable. Either "Yes" or "No". Where a value of "Very frequently" or "Sometimes" will be "Yes", and the others, "No". I'm also going to see which of these genres are listened to by at least 35% of the survey respondents. I'll mark this with a dotted black line on the graph.

```{r, r-eda-3, fig.align="center"}
df_genre_pivot_r |>
  mutate(
    value_collapsed = case_when(
      value %in% c("Never", "Rarely") ~ "No",
      value %in% c("Sometimes", "Very frequently") ~ "Yes",
    ),
    name_cleaned_ordered = fct_reorder(
      name_cleaned,
      value_collapsed == "Yes",
      .fun = sum,
      .desc = TRUE
    )
  ) |>
  ggplot(aes(x = name_cleaned_ordered, fill = value_collapsed)) +
  geom_bar(
    color = "#000000"
  ) +
  geom_hline(
    yintercept = 736 * .35,
    color = "#000000",
    linewidth = 1.5,
    linetype = 2
  ) +
  scale_fill_manual(
    values = c(
      "No" = "#FFFFFF",
      "Yes" = "#084164"
    )
  ) +
  labs(
    title = "Distribution of Music Genre Listening Frequency",
    subtitle = "Among Survey Respondents",
    fill = "Response:",
    x = "Musical Genre",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_line(color = "#000000"),
    plot.title = element_text(size = 14, face = "bold"),
    text = element_text(size = 10),
    axis.text.x.bottom = element_text(
      size = 10,
      angle = 45,
      vjust = 1,
      hjust = 1
    ),
    legend.position = "top"
  )

```

So the genres that are listened to by at least 35% of the respondents are Rock, Pop, Hip-Hop, Rap, Classical, Video Game Music, R & B, and Metal. Not bad at all. I'll drop the responses for the other genres and only focus on these eight most listened to genres.

## Python

I'm genuinely curious to see how ages play out with this. Which generation has the most anxiety on average? This data was first uploaded onto Kaggle three years ago; So I'll use the year of 2022 to create categories of generations like "Millennials", "Gen-Z", etc. We'll calculate it based on this random table Google generated:

<br>

| Generation Name             | Birth Years    | Current Age Range (as of 2022)      |
|-----------------------------|---------------|--------------------------------------|
| The Silent Generation       | 1928–1945     | 77–94 years old                      |
| Baby Boomers                | 1946–1964     | 58–76 years old                      |
| Generation X (Gen X)        | 1965–1980     | 42–57 years old                      |
| Millennials (Gen Y)         | 1981–1996     | 26–41 years old                      |
| Generation Z (Gen Z)        | 1997–2012     | 10–25 years old                      |

----

<br>

```{python, py-eda, fig.align="center"}
# Set the labels and wrap them for the plot
generation_labels = [
    "The Silent Generation",
    "Baby Boomers",
    "Gen-X",
    "Millennials",
    "Gen-Z",
]

# List comps are my favorite thing about Python so far
generation_labels_wrapped = [
    textwrap.fill(label, width=10) if len(label) > 12 else label
    for label in generation_labels
]

# Cant't calc birth year without age, so drop nulls
# then derive birth year
df_mh_music_py_eda = (
    df_mh_music_py.with_columns((2022 - pl.col("age")).alias("birth_year"))
    .with_columns(
        pl.when(pl.col("birth_year").is_between(1927, 1945, closed="both"))
        .then(pl.lit(generation_labels_wrapped[0]))
        .when(pl.col("birth_year").is_between(1946, 1964, closed="both"))
        .then(pl.lit(generation_labels_wrapped[1]))
        .when(pl.col("birth_year").is_between(1965, 1980, closed="both"))
        .then(pl.lit(generation_labels_wrapped[2]))
        .when(pl.col("birth_year").is_between(1981, 1996, closed="both"))
        .then(pl.lit(generation_labels_wrapped[3]))
        .otherwise(pl.lit(generation_labels_wrapped[4]))
        .alias("generation")
    )
    .group_by("generation")
    .agg(pl.col("anxiety").mean().alias("mean_anxiety"), pl.len().alias("n_generation"))
)

df_mh_music_py_eda = df_mh_music_py_eda.with_columns(
    pl.col("mean_anxiety")
    .map_elements(lambda x: f"{x:.2f}" if x is not None else None)
    .alias("top_label"),
    ("(n = " + pl.col("n_generation").cast(str) + ")").alias("bottom_label"),
)

(
    ggplot(df_mh_music_py_eda, aes(x="generation", y="mean_anxiety", fill="mean_anxiety"))
    + geom_bar(stat="identity", color="#000000")
    + geom_text(
        label=df_mh_music_py_eda["top_label"], nudge_y=-0.5, size=13, fontweight="bold"
    )
    + geom_text(label=df_mh_music_py_eda["bottom_label"], nudge_y=-1.1, fontstyle="italic")
    + scale_x_discrete(limits=generation_labels_wrapped)
    + scale_fill_gradient2(low="#f0cf59", high="#822801")
    + theme_minimal()
    + theme(
        plot_title=element_text(size=14, face="bold"),
        text=element_text(size=10),
        legend_position="top",
        axis_text_y=element_blank(),
        axis_title_y=element_blank(),
        axis_title_x=element_blank(),
        axis_text_x=element_text(size=10, face="demibold", color="#000000"),
    )
    + labs(
        title="Mean Self-Reported Anxiety Score",
        subtitle="Among Generational Groups",
        fill="Average Anxiety Score:",
    )
)

```

<br>

To absolutely no one's surprise, millennials and Gen-Z have the highest reported scores of anxiety. But also, as I expected, major selection bias is going on here as the sample sizes among the generations are super imbalanced. Do we really have 2 respondents from the silent generation? Sounds sus. I expected 0. It's also interesting that Millennials and Gen-Z are almost identical. We're just doomed I guess 🫠
:::
# Data Cleaning & Preprocessing

::: {.panel-tabset collapse=true}

## R

Now, we'll just grab the columns we need and make sure there's no missing values or anything like that. In R, we can make our `frequency_*` variables into factors as well. As mentioned earlier, there are four potential responses that a genre can have to answer to the question "How often do you listen to * genre?"

```{r, r-clean}
genres_to_keep <-
  paste0(
    "frequency_",
    c(
      "rock",
      "pop",
      "hip_hop",
      "rap",
      "classical",
      "video_game_music",
      "r_b",
      "metal"
    )
  )

df_mh_music_r_sub <-
  df_mh_music_r |>
  # Only select what we need
  select(anxiety, genres_to_keep) |>
  drop_na() |>
  mutate(
    across(
      starts_with("Frequency"),
      \(x) {
        factor(
          x,
          levels = c("Never", "Rarely", "Sometimes", "Very frequently")
        )
      }
    )
  )

```

## Python

In python, there isn't a direct "factor" equivalent like in R. We can make an equivalent while processing the data after we split it in the nest step. For now, I'll just select the relevant columns and make the frequency columns are casted as categorical just to be explicit. 

```{python}
import polars as pl

# Select the relevant columns
freq_cols = ["frequency_" + prefix for prefix in ["rock",
    "pop",
    "hip_hop",
    "rap",
    "classical",
    "video_game_music",
    "rb",
    "metal",
]]

df_mh_music_py_sub = df_mh_music_py.select(["anxiety"] + freq_cols).drop_nulls()

# Casting frequency columns to categorical
df_mh_music_py_sub = df_mh_music_py_sub.with_columns(
    [pl.col(col).cast(pl.Categorical) for col in freq_cols]
)

```

:::

# Train/Test Split

From this point on, I'm going to try and mirror tidymodels as closely as I can based on scikit-learn's current APIs.

::: {.panel-tabset collapse=true}

## R

Since this will be a basic analysis, I'll do a simple test/training split while keeping {rsample}'s default proportion of 75%. Meaning we'll retain 75% of the data for training and test the remaining 25%. I'll set a seed for reproducibility.

```{r, r-split}
set.seed(901622)

# Make a split w/ rsample
mh_music_r_split <-
  df_mh_music_r_sub |>
  initial_split()

df_music_r_testing <-
  mh_music_r_split |>
  testing()

df_music_r_training <-
  mh_music_r_split |>
  training()
```

## Python

We'll also do a 75% test/train split in sci-kit learn. The `sklearn.model_selection` submodule has `train_test_split` which seems to be the closest equivalent to {rasmple}'s `initial_split`.
```{python, py-split}
from sklearn.model_selection import train_test_split

# Features (frequency columns)
feats = df_mh_music_py_sub.drop("anxiety")

# Target (Anxiety score)
target = df_mh_music_py_sub.select("anxiety")

# 75% train 25% test split four dfs in total. test and trains for both features and target
df_feats_train, df_feats_test, df_target_train, df_target_test = train_test_split(
    feats, target, test_size=0.25, random_state=901622
)
```

:::


# Model Building

## Recipe Definition

::: {.panel-tabset collapse=true}

## R

Next, I'll make a recipe for the training set that will convert the frequency columns into dummies. I'm also noting that given the model I'm choosing in the next step, this probably isn't required in {parsnip}, but I wanted to make the comparison of this side-by-side with python.

```{r, r-recipe}
mh_music_recipe <-
  recipe(anxiety ~ ., df_music_r_training) |>
  step_dummy(all_nominal_predictors())
```

## Python

Scikit-learn doesn't have the "recipe" framework we can use in tidymodels, but it does have `ColumnTransformer` in its `sklearn.compose` submodule that will allow us to create the same post-processing steps.
```{python, py-recipe}
from sklearn.preprocessing import OneHotEncoder
from sklearn.compose import ColumnTransformer

frequency_categories = ["Never", "Rarely", "Sometimes", "Very frequently"]

frequency_transformer = ColumnTransformer(
    transformers=[
        (
            "freq_onehot",
            OneHotEncoder(
                categories=[frequency_categories] * len(freq_cols),
                handle_unknown="ignore",
                sparse_output=False,  # <-- use this instead of sparse=False
            ),
            freq_cols,
        )
    ],
    remainder="passthrough",
    verbose_feature_names_out=False,
)

# Convert back to polars
frequency_transformer.set_output(transform="polars")

# Fit on training set
df_feats_train_transformed = frequency_transformer.fit_transform(df_feats_train)

# Transform test set
df_feats_test_transformed = frequency_transformer.transform(df_feats_test)


```

:::

## Model Specification

::: {.panel-tabset collapse=true}

## R

I'm not trying to get too into the weeds here, but I'll do simple tuning for `cost_complexity`, `tree_depth`, and `min_n`. I'll use the "rpart" engine for proper recursion through the trees. Lastly, because our target is an anxiety score, we'll set the model to "regression". 
```{r, r-model-spec}
model_tree_r <-
  decision_tree(
    cost_complexity = tune(),
    tree_depth = tune(),
    min_n = tune()
  ) |>
  set_engine("rpart") |>
  set_mode("regression")
```

## Python

In Python, the process is similar. Using the tree module and `DecisionTreeRegressor` class. The training features and target go in the `fit` class constructor, and the testing features go into the `predict` class constructor.
```{python, py-model-spec}
from sklearn import tree
model_tree_py = tree.DecisionTreeRegressor(random_state=901622, max_depth=5)

```

:::

## Workflow Creation

::: {.panel-tabset collapse=true}

## R

```{r, r-workflow}
workflow_r <-
  workflow() |>
  add_recipe(mh_music_recipe) |>
  add_model(model_tree_r)

```

## Python

```{python, py-workflow}
grid_search = grid_search.fit(df_feats_train_transformed, df_target_train)
final_model_py = grid_search.best_estimator_
workflow_py = final_model_py.predict(df_feats_test_transformed)

```

:::

# Model Tuning

## Cross-Validation Setup
::: {.panel-tabset collapse=true}

## R

```{r, r-cross-val}
set.seed(901622)
df_music_folds_r <- vfold_cv(df_music_r_training, v = 5)
```

## Python

```{python}
from sklearn.model_selection import cross_val_score

music_folds_py = cross_val_score(
  model_tree_py,
  df_feats_train_transformed,
  df_target_train,
  cv = 5
  )

```

:::

## Tuning Grid & Selection
::: {.panel-tabset collapse=true}

## R

```{r, r-grid-tuning-1}
df_tree_grid_r <- grid_regular(
  cost_complexity(range = c(-10, -3), trans = scales::log10_trans()),
  tree_depth(range = c(2, 10)),
  min_n(range = c(2, 20)),
  levels = 5
)

df_mh_music_tune_results_r <-
  workflow_r |>
  tune_grid(
    resamples = df_music_folds_r,
    grid = df_tree_grid_r,
    metrics = metric_set(rmse, rsq)
  )

show_best(df_mh_music_tune_results_r, metric = "rmse")
best_params_r <- select_best(df_mh_music_tune_results_r, metric = "rmse")

```

When tuning these results, I actually got a warning:

```r
→ A | warning: A correlation computation is required,
but `estimate` is constant and has 0 standard deviation,
resulting in a divide by 0 error. `NA` will be returned.
There were issues with some computations   A: x5
```

Looking at my best models above, I can verify that these warnings were not referring to my best models here. So I'll continue on with the best one I have:

```{r, r-grid-tuning-2}
best_params_r
```

## Python

```{python, py-grid-tuning}
from sklearn.model_selection import GridSearchCV

param_grid = {
    "ccp_alpha": [1.000000e-10, 5.623413e-09, 3.162278e-07, 1.778279e-05, 1.000000e-03],
    "max_depth": [2, 4, 6, 8, 10],           # matches tree_depth in R
    "min_samples_leaf": [2, 6, 10, 14, 18],  # matches min_n in R
}

grid_search = GridSearchCV(
    estimator=model_tree_py,
    param_grid=param_grid,
    cv=5
)
grid_search = grid_search.fit(df_feats_train_transformed, df_target_train)

```

:::

# Final Model Tuning

::: {.panel-tabset collapse=true}

## R

```{r}
final_workflow_r <-
  workflow_r |>
  finalize_workflow(best_params_r)

final_fit_r <-
  final_workflow_r |>
  fit(df_music_r_training)
```

## Python

```{python}
final_model_py = grid_search.best_estimator_
final_model_py.fit(df_feats_train_transformed, df_target_train)
```

:::

# Model Evaluation

::: {.panel-tabset collapse=true}

## R

```{r}
df_music_predictions <-
  predict(final_fit_r, df_music_r_testing) |>
  bind_cols(df_music_r_testing)

metrics(df_music_predictions, truth = anxiety, estimate = .pred)

ggplot(df_music_predictions, aes(x = anxiety, y = .pred)) +
  geom_point(alpha = 0.5) +
  geom_abline(color = "red") +
  labs(
    x = "Actual Anxiety",
    y = "Predicted Anxiety",
    title = "Tuned Decision Tree Predictions"
  )
```

## Python

```{python}
from sklearn.metrics import mean_squared_error, mean_absolute_error, r2_score
import numpy as np

# Predict on the test set
y_pred = final_model_py.predict(df_feats_test_transformed)

# Combine predictions and actuals into a DataFrame
df_music_predictions = pl.DataFrame(
    {"anxiety": df_target_test["anxiety"], "pred": y_pred}
)

# Calculate metrics
mse = mean_squared_error(df_music_predictions["anxiety"], df_music_predictions["pred"])
rmse = np.sqrt(mse)
mae = mean_absolute_error(df_music_predictions["anxiety"], df_music_predictions["pred"])
r2 = r2_score(df_music_predictions["anxiety"], df_music_predictions["pred"])

# Print metrics
rmse, mae, r2

# Plot actual vs predicted
(
    ggplot(df_music_predictions, aes(x="anxiety", y="pred"))
    + geom_point(alpha=0.5)
    + geom_abline(color="red")
    + labs(
        x="Actual Anxiety",
        y="Predicted Anxiety",
        title="Tuned Decision Tree Predictions",
    )
)
```

:::

# Comparison & Takeaways

# Conclusion

```{python, scratch-code}
#| echo: false
#| eval: false

import numpy as np
from sklearn.model_selection import train_test_split, GridSearchCV
from sklearn.preprocessing import MinMaxScaler
from sklearn.neighbors import KNeighborsClassifier
from sklearn.metrics import accuracy_score, confusion_matrix
import plotnine as p9
```