---
title: "Mental Health and Musical Preferences"
author: "Meghan Harris"
format: html
execute:
  echo: true
  warning: false
  error: true
engine: knitr
code-fold: true
# Extension configuration
filters:
  - collapse-output
extensions:
  collapse-output:
    method: javascript
format-options:
  python:
    code-formatter: none
    path: ".venv/Scripts/python.exe"
---

# Introduction

I haven't done statistical modeling since 2021, and back then I tried using base R's `lm()` and hated it. I didn't know tidymodels existed at the time, so I assumed modeling just wasn't for me. Fast forward to 2024: I'm pivoting from data science to open source development, learning Python alongside R, and decided to revisit modeling with modern tools. This notebook compares a simple ML workflow in both R (tidymodels) and Python (scikit-learn) to:

1. Refresh my modeling skills using tidymodels (which actually makes sense to my brain)
2. Practice Python by replicating a workflow I understand in R

I'm using the [Music & Mental Health Survey Results](https://www.kaggle.com/datasets/catherinerasgaitis/mxmh-survey-results) dataset from Kaggle to predict anxiety scores from music listening habits.


# Data Import and Viewing
First, we import the data.

::: {.panel-tabset}

## R

```{r, r-data-import}
# install.packages(c("janitor", "tidyverse", "tidymodels"))
library(tidymodels)
library(tidyverse)

# Grab relative data path
data_path <-
  here::here(
    "mental_health_music",
    "data",
    "mxmh_survey_results.csv"
  )

# Import data and clean vars
df_mh_music_r <-
  read_csv(data_path) |>
  janitor::clean_names()

```

## Python

```{python, py-data-import}
import polars as pl
import janitor.polars
from pathlib import Path
from plotnine import ggplot, aes, after_stat, element_blank, element_text, geom_bar, geom_text, labs, scale_fill_gradient2, scale_x_discrete, theme, theme_minimal
import textwrap

# Python has been SUPER annoying with paths for me in quarto docs
# Maybe it's me/my machine but, meh
# Setting two possible locations for the data file
possible_paths = [
    Path("data") / "mxmh_survey_results.csv",
    Path("mental_health_music") / "data" / "mxmh_survey_results.csv",
]

for path in possible_paths:
    if path.exists():
        data_path = path
        break
else:
    raise FileNotFoundError("mxmh_survey_results.csv not found in expected locations.")

# Read in csv and clean up the names
df_mh_music_py = pl.read_csv(
    data_path,
    # This is needed because there's decimals in the
    # scoring cols where integer was inferred
    infer_schema_length=1000,
).clean_names(remove_special=True)

```

:::

Then, do a little skimming with glimpse...

::: {.panel-tabset collapse=true}

## R

```{r, r-data-glimpse}
#| output-fold: true
library(dplyr)

# Take a peek
df_mh_music_r |>
  glimpse()

```

## Python

```{python, py-data-glimpse}
# | output-fold: true
# Take a peek
df_mh_music_py.glimpse()
```

:::

# Data Exploration

Looking at the data, I'm interested in predicting "anxiety score" from musical listening patterns with respect to genre frequencies.
We'll need the `anxiety`, `frequency_*` and maybe the `hours_per_day` variables, so let's visualize what we have here.

For the anxiety score interpretation on a scale from 0 to 10, respondents where asked "how often do you experience this?":
  
  * 0 - I do not experience this.

  * 10 - I experience this regularly, constantly/or to an extreme.

::: {.panel-tabset collapse=true}

## R

So, of no surprise to me, this data is right-skewed with a concentration sitting around a score from 7 - 8.
The density curve also shows limited variability at lower anxiety levels ("flattening" of the curve), which may affect model performance in predicting across the full range of the outcome.

```{r, r-eda, fig.align="center"}
df_mh_music_r |>
  ggplot(aes(x = anxiety)) +
  geom_histogram(
    aes(y = after_stat(density), fill = after_stat(count)),
    bins = 11,
    color = "#000000"
  ) +
  geom_density(color = "#000000", linewidth = 3) +
  geom_density(color = "#a873d1", linewidth = 1) +
  scale_fill_gradient2(
    low = "#ffffff",
    high = "#250647"
  ) +
  labs(
    title = "Distribution of Self-Reported Anxiety Scores",
    subtitle = "Among Survey Respondents",
    fill = "Total Responses:",
    x = "Reported Anxiety Score",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    text = element_text(size = 10),
    legend.position = "top"
  )
```

## Python

I'm genuinely curious to see how ages play out with this. Which generation has the most anxiety on average? This data was first uploaded onto Kaggle three years ago; So I'll use the year of 2022 to create categories of generations like "Millennials", "Gen-Z", etc. We'll calculate it based on this random table Google generated:

<br>

| Generation Name             | Birth Years    | Current Age Range (as of 2022)      |
|-----------------------------|---------------|--------------------------------------|
| The Silent Generation       | 1928–1945     | 77–94 years old                      |
| Baby Boomers                | 1946–1964     | 58–76 years old                      |
| Generation X (Gen X)        | 1965–1980     | 42–57 years old                      |
| Millennials (Gen Y)         | 1981–1996     | 26–41 years old                      |
| Generation Z (Gen Z)        | 1997–2012     | 10–25 years old                      |

----

<br>

```{python, py-eda , fig.align="center"}
# Set the labels and wrap them for the plot
generation_labels = [
    "The Silent Generation",
    "Baby Boomers",
    "Gen-X",
    "Millennials",
    "Gen-Z",
]

# List comps are my favorite thing about Python so far
generation_labels_wrapped = [
    textwrap.fill(label, width=10) if len(label) > 12 else label
    for label in generation_labels
]

# Cant't calc birth year without age, so drop nulls
# then derive birth year
df_mh_music_py = (
    df_mh_music_py.with_columns((2022 - pl.col("age")).alias("birth_year"))
    .with_columns(
        pl.when(pl.col("birth_year").is_between(1927, 1945, closed="both"))
        .then(pl.lit(generation_labels_wrapped[0]))
        .when(pl.col("birth_year").is_between(1946, 1964, closed="both"))
        .then(pl.lit(generation_labels_wrapped[1]))
        .when(pl.col("birth_year").is_between(1965, 1980, closed="both"))
        .then(pl.lit(generation_labels_wrapped[2]))
        .when(pl.col("birth_year").is_between(1981, 1996, closed="both"))
        .then(pl.lit(generation_labels_wrapped[3]))
        .otherwise(pl.lit(generation_labels_wrapped[4]))
        .alias("generation")
    )
    .group_by("generation")
    .agg(pl.col("anxiety").mean().alias("mean_anxiety"), pl.len().alias("n_generation"))
)

df_mh_music_py = df_mh_music_py.with_columns(
    pl.col("mean_anxiety")
    .map_elements(lambda x: f"{x:.2f}" if x is not None else None)
    .alias("top_label"),
    ("(n = " + pl.col("n_generation").cast(str) + ")").alias("bottom_label"),
)


(
    ggplot(df_mh_music_py, aes(x="generation", y="mean_anxiety", fill="mean_anxiety"))
    + geom_bar(stat="identity", color="#000000")
    + geom_text(
        label=df_mh_music_py["top_label"], nudge_y=-0.5, size=13, fontweight="bold"
    )
    + geom_text(label=df_mh_music_py["bottom_label"], nudge_y=-1.1, fontstyle="italic")
    + scale_x_discrete(limits=generation_labels_wrapped)
    + scale_fill_gradient2(low="#f0cf59", high="#822801")
    + theme_minimal()
    + theme(
        plot_title=element_text(size=14, face="bold"),
        text=element_text(size=10),
        legend_position="top",
        axis_text_y=element_blank(),
        axis_title_y=element_blank(),
        axis_title_x=element_blank(),
        axis_text_x=element_text(size=10, face="demibold", color="#000000"),
    )
    + labs(
        title="Mean Self-Reported Anxiety Score",
        subtitle="Among Generational Groups",
        fill="Average Anxiety Score:",
    )
)

```

<br>

To absolutely no one's surprise, millennials and Gen-Z have the highest reported scores of anxiety. But also, as I expected, major selection bias is going on here as the sample sizes among the generations are super imbalanced. Do we really have 2 respondents from the silent generation? Sounds sus. I expected 0. It's also interesting that Millennials and Gen-Z are almost identical. We're just doomed I guess 🫠
:::
# Data Cleaning 

::: {.panel-tabset collapse=true}

## R

```{r}

```

## Python

```{python}


```

:::

# Train/Test Split

::: {.panel-tabset collapse=true}

## R

```{r}

```

## Python

```{python}


```

:::


# Model Building

## Recipe Definition

::: {.panel-tabset collapse=true}

## R

```{r}

```

## Python

```{python}


```

:::

## Model Specification

::: {.panel-tabset collapse=true}

## R

```{r}

```

## Python

```{python}


```

:::

## Workflow Creation

::: {.panel-tabset collapse=true}

## R

```{r}

```

## Python

```{python}


```

:::

# Model Tuning

## Cross-Validation Setup
::: {.panel-tabset collapse=true}

## R

```{r}

```

## Python

```{python}


```

:::

## Tuning Grid & Selection
::: {.panel-tabset collapse=true}

## R

```{r}

```

## Python

```{python}


```

:::

# Final Model Tuning

::: {.panel-tabset collapse=true}

## R

```{r}

```

## Python

```{python}


```

:::

# Model Evaluation

::: {.panel-tabset collapse=true}

## R

```{r}

```

## Python

```{python}


```

:::

# Comparison & Takeaways

# Conclusion

```{python, scratch-code}
#| echo: false
#| eval: false

import numpy as np
from sklearn.model_selection import train_test_split, GridSearchCV
from sklearn.preprocessing import MinMaxScaler
from sklearn.neighbors import KNeighborsClassifier
from sklearn.metrics import accuracy_score, confusion_matrix
import plotnine as p9
```